\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgment}
This thesis would not have been made possible without the continuous support and help from my Ph.D. supervisor François Yvon and my Ph.D. co-supervisor Josep Maria Crego. They spent countless hours discussing research with me, helping me with paper writing, and were very supportive throughout my Ph.D. The outstanding critical thinking of François will influence me for the rest of my research career. I have also learned a lot from Josep's problem-solving skills. I consider myself incredibly fortunate to have been their student. My second thank goes to all the members of the jury committee, Alexander Fraser, Marine Carpuat, Sennrich Rico, who gave me many insights during the defense and also helped me correct many errors in my thesis.

Then, I would like to thank my parents and my little brother. Without their emotional support, I could not have gone this far in my research career. I want to thank also my long-time girlfriend, who also is doing a Ph.D. thesis, for having been very supportive during my Ph.D. and for her helpful outside perspectives.

Next, I would like to thank my colleagues in Systran: Guillaume Klein, Natalia Segal, Mickael Mani, and Elise Michon for many running sessions. I will always cherish the race Paris-Versailles 2018. My thank also goes to Jean Senellart, CEO of Systran, for helping to create this collaborative project between Systran and LISN. 

Besides, I would like to thank many friends at LISN,  Aina Gari-Soler, François Buet, Alban Petit, Aman Berhe, Jitao Xu, Paul Lerner, Marc Benzahra, Shu Okabe and Anh Khoa Ngo Ho. I always cherish our coffee breaks.

This work was generously granted access to the HPC resources of [TGCC/CINES/IDRIS] under the allocation 2020- [AD011011270] made by GENCI (Grand Equipement National de Calcul Intensif).
